---
title: "A3"
author: "Maiko Hata"
format: html
editor: visual
---

# **EDLD 654, Fall 2025: Assignment 3**

The purpose of this assignment is to get you working with the `caret` package and `ranger` and `gbm` engine to fit Bagged Trees, Random Forests, and Gradient Boosting Trees for predicting a continuous or a categorical outcome. You will use the same datasets and recipes from the previous two assigments. Assignment 1.

There are alternative ways to submit your assignment depending on your preference.

1.  You can Copy/Edit this notebook and complete it with your responses. Then, you can save and run the completed Kaggle notebook and submit the link through Canvas. If you keep your notebook Private, do not forget to share it with 'Lina Shanley PhD' so I can access it.

2.  You copy/paste the questions and download the datasets from the notebook to your computer. Then, complete the assignment as an R markdown document. Then, you can knit the R Markdown document to a PDF and submit both the .Rmd and PDF files by uploading them on Canvas.

3.  You knit the R Markdown document to an HTML document and host it on your website/blog or any publicly available platform. Then, you can submit the .Rmd file by uploading it on Canvas and putting the link for the HTML document as a comment.

4.  If you have a GitHub repo and store all your work for this class in a GitHub repo, you can create a folder for this assignment in that repo and put the.Rmd file and PDF document under a specific folder. Then, you can submit the link for the GitHub repo on Canvas.

To receive full credit, you must complete the following tasks. Please make sure that all the R code you wrote for completing these tasks and any associated output are explicitly printed in your submitted document. If the task asks you to submit the data files you created, please upload these datasets along with your submission.

If you have any questions, please do not hesitate to reach out to me.

## **Task 1: Predicting a Categorical Outcome using Bagged Trees, Random Forests, and Gradient Boosting Trees**

### **Description**

For this assignment you will work with the twitter dataset and build models to predict the sentiment of a given tweet. We will use the same dataset from the previous two assignments. Processed tweet data is available as input data in this notebook. Please use the tweet data provided as input in this notebook, so there is consistency of code across people.

First, import the tweet data into the R environment (**../input/assignment1tweet-final/tweet_final.csv**). You can give any name to this data object. Print the structure of this data object using the `str` function.

```{r}
tweet <- read.csv('/Users/maiko/Desktop/EDLD654_ML/data/tweet_final.csv')
```

```{r}
require(tidyverse)
require(recipes)

blueprint_tweet <- recipe(x  = tweet,
                          vars  = colnames(tweet),
                          roles = c('outcome',rep('predictor',772))) %>%
  step_dummy('month',one_hot=TRUE) %>% 
  step_harmonic('day',frequency=1,cycle_size=7, role='predictor') %>%
  step_harmonic('date',frequency=1,cycle_size=31,role='predictor') %>%
  step_harmonic('hour',frequency=1,cycle_size=24,role='predictor') %>%
  step_normalize(paste0('V',1:768)) %>%
  step_normalize(c('day_sin_1','day_cos_1',
                   'date_sin_1','date_cos_1',
                   'hour_sin_1','hour_cos_1')) %>%
  step_num2factor(sentiment,
                  transform = function(x) x + 1,
                  levels=c('Negative','Positive'))
```

```{r}
View(blueprint_tweet %>% prep() %>% summary)

    # There are 777 predictors in the processed data
    # 768 sentence embeddings + 3 dummy variables for month + 6 harmonic terms (2 variables per harmonic variable)
```

### **Tasks**

**Task 1.1** Split the original data into two subsets: training and test. Let the training data have the 80% of cases and the test data have the 20% of the cases. Set the seed to **10312022** for any random sampling process before splitting data.

```{r}
library(ranger)
library(pROC)
library(caret)
```

```{r}
set.seed(10312022) 

loc_tweet <- sample(1:nrow(tweet), round(nrow(tweet) * 0.8))
tweet_tr  <- tweet[loc_tweet, ]
tweet_te  <- tweet[-loc_tweet, ]

dim(tweet_tr)
dim(tweet_te)
```

#### **Bagged Trees**

**Task 1.2** Use the `caret::train()` function and `ranger` engine to train a model with 10-fold cross-validation for predicting the probability of sentiment being positive using a Bagged Trees model with 500 trees. Set the `max.dept` argument to 60 and the `min.node.size` argument to 30. Note that you are supposed to use all predictors in the processed data for Bagged Trees.

```{r}
prep_tweet <- blueprint_tweet %>% prep()

tweet_tr_processed <- juice(prep_tweet)
tweet_te_processed <- bake(prep_tweet, tweet_te)
```

```{r}
set.seed(10312022)

bagged_tweet <- ranger(
  sentiment ~ .,
  data = tweet_tr_processed,
  num.trees = 500,
  mtry = ncol(tweet_tr_processed) - 1,   # use all predictors
  min.node.size = 30,
  max.depth = 60,
  probability = TRUE,
  importance = "none"
)

bagged_tweet
```

**Task 1.3** Evaluate the performance of Bagged Trees model on the test set by computing AUC, TPR, TNR, FPR, and Precision.

```{r}
# Step 1: Predict probabilities on the test set
bagged_probs <- predict(bagged_tweet, tweet_te_processed)$predictions[,"Positive"]
head(bagged_probs)

```

```{r}
# Step 2: Convert probabilities to class predictions
bagged_class <- ifelse(bagged_probs > 0.5, "Positive", "Negative") %>% factor()
```

```{r}
library(pROC)

# Step 3: Compute AUC
bagged_auc <- roc(response = tweet_te_processed$sentiment,
                  predictor = bagged_probs,
                  levels = c("Negative", "Positive"))

bagged_auc_value <- auc(bagged_auc)
bagged_auc_value
```

```{r}
# Step 4: Confusion matrix
cm <- table(Predicted = bagged_class,
            Actual    = tweet_te_processed$sentiment)
cm
```

```{r}
TP <- cm["Positive","Positive"]
TN <- cm["Negative","Negative"]
FP <- cm["Positive","Negative"]
FN <- cm["Negative","Positive"]
```

#### **Random Forests**

**Task 1.4.** Use the `caret::train()` function and `ranger` engine to train a model with 10-fold cross-validation for predicting the probability of sentiment being positive using a Random Forest model with 500 trees. Set the `max.dept` argument to 60 and the `min.node.size` argument to 30. Also, set the number of predictors to consider to 500 for each tree while growing a random forest.

```{r}
library(caret)
library(ranger)

set.seed(10312022)

ctrl_rf <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)
```

```{r}
grid_rf <- expand.grid(
  mtry = 500,          # assignment requirement
  splitrule = "gini",  # classification
  min.node.size = 30
)
```

```{r}
# Step 3: Train Random Forest
rf_model <- train(
  sentiment ~ .,
  data = tweet_tr_processed,
  method = "ranger",
  trControl = ctrl_rf,
  tuneGrid = grid_rf,
  num.trees = 500,
  max.depth = 60
)

rf_model
```

**Task 1.5** Evaluate the performance of the Random Forest model on the test set by computing AUC, TPR, TNR, FPR, and Precision.

```{r}
rf_probs <- predict(rf_model, tweet_te_processed, type = "prob")[,"Positive"]
head(rf_probs)
```

```{r}
rf_class <- ifelse(rf_probs > 0.5, "Positive", "Negative") %>% factor()
```

```{r}
rf_auc <- roc(
  response = tweet_te_processed$sentiment,
  predictor = rf_probs,
  levels = c("Negative", "Positive")
)

auc(rf_auc)

```

```{r}
cm_rf <- table(
  Predicted = rf_class,
  Actual = tweet_te_processed$sentiment
)

cm_rf

```

```{r}
TP <- cm_rf["Positive","Positive"]
TN <- cm_rf["Negative","Negative"]
FP <- cm_rf["Positive","Negative"]
FN <- cm_rf["Negative","Positive"]

TPR <- TP / (TP + FN)
TNR <- TN / (TN + FP)
FPR <- FP / (FP + TN)
Precision <- TP / (TP + FP)

TPR
TNR
FPR
Precision

```

#### **Gradient Boosting Trees**

**Task 1.6.** Use the `caret::train()` function and `gbm` engine to train a model with 10-fold cross-validation for predicting the probability of sentiment being positive using a Gradient Boosting Tree model with 250 iterations (trees). Set the `interaction.depth` argument to 9, `shrinkage` argument to 0.05, and `n.minobsinnode` argument to 30. Also, set the `bag.fraction` argument to 0.5.

```{r}
library(gbm)
```

```{r}
ctrl_gbm <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)
```

```{r}
grid_gbm <- expand.grid(
  n.trees = 250,
  interaction.depth = 9,
  shrinkage = 0.05,
  n.minobsinnode = 30
)

```

```{r}
set.seed(10312022)

gbm_model <- train(
  sentiment ~ .,
  data = tweet_tr_processed,
  method = "gbm",
  trControl = ctrl_gbm,
  tuneGrid = grid_gbm,
  bag.fraction = 0.5,
  verbose = FALSE
)

```

```{r}
install.packages("beepr")
library(beepr)

beep_on_done <- function(expr) {
  result <- eval(expr)
  beepr::beep(1)
  return(result)
}

```

```{r}
gbm_model
```

**Task 1.7** Evaluate the performance of the Gradient Boosting model on the test set by computing AUC, TPR, TNR, FPR, and Precision.

```{r}
gbm_probs <- predict(gbm_model, tweet_te_processed, type = "prob")[,"Positive"]
head(gbm_probs)

```

```{r}
gbm_class <- ifelse(gbm_probs > 0.5, "Positive", "Negative") %>% factor()

```

```{r}
gbm_auc <- roc(
  response = tweet_te_processed$sentiment,
  predictor = gbm_probs,
  levels = c("Negative","Positive")
)

auc(gbm_auc)

```

```{r}
cm_gbm <- table(
  Predicted = gbm_class,
  Actual = tweet_te_processed$sentiment
)

cm_gbm

```

```{r}
TP <- cm_gbm["Positive","Positive"]
TN <- cm_gbm["Negative","Negative"]
FP <- cm_gbm["Positive","Negative"]
FN <- cm_gbm["Negative","Positive"]

TPR <- TP / (TP + FN)
TNR <- TN / (TN + FP)
FPR <- FP / (FP + TN)
Precision <- TP / (TP + FP)

TPR
TNR
FPR
Precision

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
