---
title: "autism_project"
author: "Maiko Hata"
format: html
editor: visual
---

```{r}
library(dplyr)
library(finalfit)
library(lubridate)
library(reticulate)
library(finalfit)
library(stringr)
library(recipes)
library(ggplot2)
library(glmnet)
library(caret)
library(vip)
```

```{r}
setwd("~/Desktop/EDLD654_ML")
```

### Following the steps on Kaggle **Assignment 1**

Task 1.1. Read data

```{r}
autism <- read.csv('/Users/maiko/Desktop/EDLD654_ML/data/Autism_Screening.csv', header=TRUE)
```

```{r}
str(autism)
```

Task 1.2. through Task 1.8. Recode cyclical data, text data – N/A\
Task 1.3. Recode text data into binary variable\
—\> start here by recoding gender, jundice, autism, class/ASD as binary?

```{r}
autism$gender_binary <- ifelse(autism$gender == "m", 1, 0)
autism$jundice_binary <- ifelse(autism$jundice == "yes", 1, 0)
autism$Class.ASD_binary <- ifelse(autism$Class.ASD == "YES", 1, 0)

# It made a new column, will need to delete original later 
```

```{r}
head(autism[c("gender", "gender_binary")])
```

```{r}
table(autism$gender_binary)
```

Task 2.2 Using the `ff_glimpse()` function from the `finalfit` package, provide a snapshot of missingness in this dataset. This function also returns the number of levels for categorical variables. If there is any variable with large amount of missingness (e.g. more than 75%), remove this variable from the dataset.

```{r}
ff_glimpse(autism)
```

```{r}
ff_glimpse(autism[, 1:20])
```

```{r}
autism <- autism[, colMeans(is.na(autism)) <= 0.75]
```

```{r}
str(autism)
```

✅ 1️⃣ Count how many “?” entries exist in `ethnicity`

```{r}
sum(autism$ethnicity == "?", na.rm = TRUE)
```

✅ 2️⃣ Check total rows (for percentage)

```{r}
nrow(autism)
```

✅ 3️⃣ Compute percentage of missing (“?”) entries (--\> I decided it's good enough, as the example from class had 75% threshold)

```{r}
mean(autism$ethnicity == "?", na.rm = TRUE) * 100
```

Yep—those quotes are literally part of the values (e.g., `'United States'` vs `Bahamas`). That comes from inconsistent quoting in the source or how it was imported, and it will bite you later because they’ll become different factor levels. If you’ve already loaded it, clean the column

```{r}
library(dplyr)
library(stringr)

autism <- autism %>%
  mutate(
    contry_of_res = str_trim(contry_of_res),
    # remove leading/trailing single or double quotes
    contry_of_res = str_replace_all(contry_of_res, "^[\"']+|[\"']+$", ""),
    # collapse internal extra spaces
    contry_of_res = str_squish(contry_of_res),
    # optional: normalize common variants
    contry_of_res = recode(contry_of_res,
      "USA" = "United States", "U.S." = "United States",
      "United States of America" = "United States"
    )
  )
```

```{r}
autism %>% count(contry_of_res, sort = TRUE)
```

Fixing the "ethnicity" column the same way as above to remove ' " etc.

```{r}
autism <- autism %>%
  mutate(
    ethnicity = str_trim(ethnicity),
    ethnicity = str_replace_all(ethnicity, "^[\"']+|[\"']+$", ""),
    ethnicity = str_squish(ethnicity)
    )
```

✅ If you’re done using the original columns\
If you’ve already converted variables like `gender`, `jaundice`, `family_mem_with_asd`, etc. into `_binary` versions (0/1) and those originals were categorical or text,\
then yes — you can safely remove them to avoid redundancy before modeling.

⚠️ If you’re still exploring or not sure yet\
Keep both for now! You can always drop the original columns later once you’ve confirmed that: 1) the `_binary` variables are correctly coded (0 for No, 1 for Yes, etc.) 2) and your model runs fine without the originals.

ASD Diagnosis Data Analysis and Prediction (<https://www.kaggle.com/code/devraai/asd-diagnosis-data-analysis-and-prediction-model>)\
— Data Cleaning —

🧩 In your **autism dataset** – If `age` is already text (e.g., `"18"`, `"2.5"`, `"?"`, `"n/a"`), then you don’t need `format()` — just convert directly:

```{r}
autism$age <- as.numeric(autism$age)
```

```{r}
sum(is.na(autism$age))
```

```{r}
# Checking to make sure "result" column is numeric already 
is.numeric(autism$result)
```

From <https://www.kaggle.com/code/devraai/asd-diagnosis-data-analysis-and-prediction-model>:

```{r}
# Distribution of ASD Class

ggplot(autism, aes(x = Class.ASD, fill = Class.ASD)) +
  geom_bar() +
  scale_fill_manual(values = c("NO" = "#69b3a2", "YES" = "#404080")) +
  labs(title = "Distribution of Autism Class",
       x = "Autism Class", y = "Count") +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  )
```

```{r}
ggplot(autism, aes(x = age)) +
  geom_histogram(binwidth = 0.5) +
  labs(title = "Age Distribution",
       x = "Age",
       y = "Count") +
  theme_minimal()
```

```{r}
ggplot(autism, aes(x = age)) +
  geom_histogram(binwidth = 0.5) +
  scale_x_continuous(limits = c(0, max(autism$age, na.rm = TRUE) + 10)) +
  labs(title = "Age Distribution",
       x = "Age",
       y = "Count") +
  theme_minimal()
```

```{r}
# Trying to remove that weird error/outliner age around 380 years old 
summary(autism$age)
```

```{r}
autism <- autism %>% 
  filter(age <= 100)
```

```{r}
# Trying again to see if the 383 got removed -- and it's much better! 🙌
ggplot(autism, aes(x = age)) +
  geom_histogram(binwidth = 0.5) +
  labs(title = "Age Distribution",
       x = "Age",
       y = "Count") +
  theme_minimal()
```

```{r}
autism <- autism %>% 
  filter(age <= 100)

ggplot(autism, aes(x = age)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue") +
  labs(title = "Age Distribution",
       x = "Age",
       y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
# Because the column "result" was numeric, I should be able to plot it hte same way: 
ggplot(autism, aes(x = result)) +
  geom_histogram(binwidth = 0.5) +
  labs(title = "Result Distribution",
       x = "Result",
       y = "Count") +
  theme_minimal()
```

```{r}
ggplot(autism, aes(x = result)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue") +
  labs(title = "Screener Results",
       x = "Score",
       y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

**Assignment 1: Task 2.3** Most of the variables in this dataset are categorical, and particularly a binary variable with a Yes and No response. Check the frequency of unique values for all categorical variables. If there is any inconsistency (e.g., Yes is coded as both 'y' and 'Y') for any of these variables in terms of how values are coded, fix them. Also, check the distribution of numeric variables and make sure there is no anomaly.

```{r}
names(autism)
```

```{r}
cat_vars <- c(
  "gender", 
  "ethnicity", 
  "jundice",       # use your dataset’s spelling
  "austim",        # same spelling as in your data
  "contry_of_res", # same spelling as in your data
  "Class.ASD"
)

num_vars <- c(
  "age"
)
### result is OUTCOME, so we exclude it ###

binary_vars <- c(
  "gender_binary",
  "jundice_binary",
  "Class.ASD_binary"
)
```

```{r}
list(
  categorical = cat_vars,
  numerical = num_vars,
  binary = binary_vars
)
```

```{r}
invisible(lapply(cat_vars, function(v) {
  cat("\n---", v, "---\n")
  print(table(autism[[v]], useNA = "ifany"))
}))
```

```{r}
summary(autism[num_vars])
```

```{r}
ggplot(autism, aes(x = ethnicity)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Race/Ethnicity",
       x = "Race/Ethnicity",
       y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
ggplot(autism, aes(x = ethnicity)) +
  geom_bar(fill = "steelblue") +
  geom_text(aes(label = after_stat(count)),
            stat = "count",
            vjust = -0.5, size = 3.5) +
  labs(title = "Race/Ethnicity",
       x = "Race/Ethnicity",
       y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

**Assignment 1: Task 2.3** Most of the variables in this dataset are categorical, and particularly a binary variable with a Yes and No response. Check the frequency of unique values for all categorical variables. If there is any inconsistency (e.g., Yes is coded as both 'y' and 'Y') for any of these variables in terms of how values are coded, fix them. Also, check the distribution of numeric variables and make sure there is no anomaly.

---\> There is no inconsistencies (I already checked and changed it (like countries to have no " ").

---\> I already checked for numerical anomaly by running summary(autism\$age) and filtered it out with autism \<- autism %\>% filter(age \<= 100)

### ✅ Step 3 — Build the modeling dataframe

why: we drop the original text columns you already replaced (gender/jundice/Class.ASD) and keep only what we’ll transform.

```{r}
autism_mod <- autism |>
  transmute(
    # outcome
    result,
    # numeric predictors
    age,
    # keep these categoricals to one-hot later
    ethnicity, contry_of_res, austim,
    # keep your existing 0/1 binaries as numeric predictors
    gender_binary, jundice_binary, Class.ASD_binary
  )
```

### ✅ 4) choose numeric vars for polynomial expansion

why: you only have age as a numeric predictor; we’ll create degree-3 raw polynomial terms for it.

```{r}
num_poly_ok <- "age"
```

### ✅ 5) define the recipe (steps only; no training yet)

### chunk 1 — recipe skeleton (declare outcome and predictors implicitly)

why: tells recipes that result is the numeric outcome; everything else in autism is a predictor.

```{r}
rec <- recipe(result ~ ., data = autism)
```

### chunk 2 — steps (1)–(4): na indicators, zero-variance, impute numeric, impute categorical

```{r}
rec <- rec |>
  step_indicate_na(all_predictors()) |>        # (1)
  step_zv(all_predictors()) |>                 # (2) catch single-level nominals like age_desc
  step_impute_mean(all_numeric_predictors()) |># (3)
  step_impute_mode(all_nominal_predictors())   # (4)
```

### chunk 3 — steps (5)–(7): polynomial(age, deg=3), standardize, one-hot dummies

```{r}
rec <- rec |>
  step_poly(age, degree = 3, options = list(raw = TRUE)) |>  # (5)
  step_normalize(all_numeric_predictors()) |>                # (6)
  step_dummy(all_nominal_predictors(), one_hot = TRUE)       # (7)
```

### chunk 4 — prep (train the recipe)

why: learns means, modes, and scaling parameters from your data.

```{r}
### Redid the chunks 2 and 3 because "You hit a known quirk of this dataset: it has a column age_desc that’s always '18 and more' (only one level). step_dummy() chokes on single-level factors. Update chunks 2–3 to remove zero-variance columns for all predictors (not just numeric):" which makes sense. 
rec_prep <- prep(rec, training = autism, verbose = TRUE)
```

### chunk 5 — blueprint

```{r}
rec_prep
tidy(rec_prep)
```

### Bake it!

```{r}
autism_xf <- bake(rec_prep, new_data = NULL)
```

```{r}
autism_final <- autism_xf |>
  dplyr::rename(score = result) |>
  dplyr::relocate(score)
```

```{r}
View(autism_final)
```

```{r}
### 701 rows, 112 variables(features) ### 
dim(autism_final)
names(autism_final)[1:20]
```

Week 5 readings (Hands on ML ch. 6)

-   Ordinary Least Squares (OLS) assumptions for linear regression (Hands On Machine Learning with R, Ch. 4 & 6):

    -   Linear relationship

    -   There are more observations (n) than features (p)

    -   No or little multicollinearity.

    **Why it matters**

-   OLS is the foundation for many regression models (linear, multiple, and even as a base for regularized ones like ridge and lasso). When you understand the OLS object and these functions, you can interpret your models and check assumptions much more easily.

```{r}
nrow(autism)
ncol(autism)
```

## ADD this part to the slides- how I decided to keep the "duplicate" data

STILL CLEANING UP AND FOUND DUPLICATES. I NEED TO LOOK AT THEM TO SEE IF IT'S DUPLICATES, OR 2 DIFFERENT PEOPLE ENTERING THE SAME DATA.

```{r}
library(dplyr)
dup_rows <- autism_final |> 
  mutate(row_id = row_number()) |> 
  filter(duplicated(across(-row_id)) | duplicated(across(-row_id), fromLast = TRUE)) |> 
  arrange_all()
dup_rows
```

## Model 1: Linear Regression without Regularization

Kaggle Notebook "Building a Ridge Regression Model\
Step 2: Train/Test split

"In order to obtain a realistic measure of model performance, we will split the data into two subsamples: training and test datasets. Due to the relatively small sample size, I will use a 90-10 split (typically a 80-20 or 70-30 split is used). The smaller test dataset will be used as a final hold-out set, and training dataset will be used to build the model."

```{r}
set.seed(20251102)

n <- nrow(autism_final)
loc <- sample(seq_len(n), size = round(0.9 * n))
autism_tr <- autism_final[loc, ]
autism_te <- autism_final[-loc, ]

length(intersect(rownames(autism_tr), rownames(autism_te)))  # should now be 0
```

```{r}
set.seed(20251102)

n   <- nrow(autism_final)
loc <- sample(seq_len(n), size = round(0.9 * n))  # train indices

autism_tr <- autism_final[loc, ]
autism_te <- autism_final[-loc, ]

# True disjointness check:
length(intersect(loc, setdiff(seq_len(n), loc)))  # should be 0

```

```{r}
lm_model <- lm(score ~ ., data = autism_tr)
```

```{r}
# BECAUSE THE A1_Score so forth were included, it predicted PERFECTLY OF COURSE that makes sense SO LATER I DROPPED THE A1_Score so forth, and also ASD.Class_Binary later
pred  <- predict(lm_model, newdata = autism_te)

R2    <- cor(pred, autism_te$score)^2
MAE   <- mean(abs(pred - autism_te$score))
RMSE  <- sqrt(mean((pred - autism_te$score)^2))  # ← missing ) added here

cat("Test R²:", round(R2, 3),
    "\nTest MAE:", round(MAE, 3),
    "\nTest RMSE:", round(RMSE, 3), "\n")

```

```{r}
library(dplyr)

# Keep score + all predictors except the A1_Score..A10_Score items
autism_tr_m <- autism_tr %>%
  select(score, -matches("^A(10|[1-9])_Score$"))

autism_te_m <- autism_te %>%
  select(score, -matches("^A(10|[1-9])_Score$"))

lm_model <- lm(score ~ ., data = autism_tr_m)

pred  <- predict(lm_model, newdata = autism_te_m)
R2    <- cor(pred, autism_te_m$score)^2
MAE   <- mean(abs(pred - autism_te_m$score))
RMSE  <- sqrt(mean((pred - autism_te_m$score)^2))

cat("Test R²:", round(R2, 3),
    "\nTest MAE:", round(MAE, 3),
    "\nTest RMSE:", round(RMSE, 3), "\n")

```

```{r}
summary(pred)
sum(is.na(pred))

```

```{r}
length(coef(lm_model))
```

```{r}
# Baseline mean-only predictions
mean_pred <- mean(autism_tr$score)
pred_base <- rep(mean_pred, nrow(autism_te))

# Compute metrics
R2_base   <- cor(pred_base, autism_te$score)^2
MAE_base  <- mean(abs(pred_base - autism_te$score))
RMSE_base <- sqrt(mean((pred_base - autism_te$score)^2))

```

### Why this happened

Once you removed the A1_Score–A10_Score columns (which directly determined the total), there wasn’t enough remaining signal in the other predictors for `lm()` to estimate meaningful slopes.

That’s why: R² = NA → no variation in predictions, MAE and RMSE = the average deviation of true scores from the mean (≈ how far actual scores differ from 4.87).

This is actually a **useful baseline** — it’s your “mean-only” model. It shows the model’s starting point: how well the simplest possible predictor (just the average score) performs.

## Model 1: Ridge Regression Model

```{r}
set.seed(20251102)

n <- nrow(autism_final)
loc <- sample(seq_len(n), size = round(0.9 * n))

autism_tr <- autism_final[loc, ]
autism_te <- autism_final[-loc, ]
```

```{r}
# ✅ Chunk 1 – Define the recipe (blueprint)
library(recipes)

blueprint <- recipe(score ~ ., data = autism_tr) %>%
  step_rm(matches("^A(10|[1-9])_Score$"), Class.ASD_binary) %>%  # drop leak-prone vars
  step_zv(all_numeric()) %>%
  step_nzv(all_numeric()) %>%
  step_normalize(all_numeric_predictors())
```

```{r}
# ✅ Chunk 2 – Create folds and cross-validation control

autism_tr <- autism_tr[sample(nrow(autism_tr)), ]
folds <- cut(seq_len(nrow(autism_tr)), breaks = 10, labels = FALSE)
my.indices <- vector("list", 10)
for (i in 1:10) my.indices[[i]] <- which(folds != i)
cv <- trainControl(method = "cv", index = my.indices)
```

```{r}
#✅ Chunk 3 – Create tuning grid and train ridge model
# Ridge grid (alpha = 0)
grid <- data.frame(alpha = 0, lambda = seq(0.01, 3, 0.01))

ridge <- caret::train(
  blueprint,
  data      = autism_tr,
  method    = "glmnet",
  trControl = cv,
  tuneGrid  = grid
)
```

```{r}
# ✅ Chunk 4 – Inspect and plot results
plot(ridge)
ridge$results
ridge$bestTune

coefs <- coef(ridge$finalModel, ridge$bestTune$lambda)
head(as.matrix(coefs[order(abs(coefs), decreasing = TRUE)[-1], ]), 10)

```

I REALIZED, AFTER THE FITTING, that Class.ASD_binary, Class.ASD_NO, Class.ASD_YES were still there predicting away based on the TOTAL SCORE (DUH) so I removed them.

```{r}
# ✅ Chunk 5 – Check top predictors
library(vip)
vip(ridge, num_features = 10, geom = "point") + theme_bw()
```

```{r}
# ✅ Chunk 6 – Visualize feature importance
predict_te_ridge <- predict(ridge, newdata = autism_te)

rsq_te  <- cor(autism_te$score, predict_te_ridge, use = "complete.obs")^2
mae_te  <- mean(abs(autism_te$score - predict_te_ridge))
rmse_te <- sqrt(mean((autism_te$score - predict_te_ridge)^2))

cat("Ridge test R²:", round(rsq_te, 3),
    "\nMAE:", round(mae_te, 3),
    "\nRMSE:", round(rmse_te, 3), "\n")
```

```{r}
vip(ridge, num_features = 10, geom = "point") + theme_bw()
```

```{r}
library(dplyr)
library(tibble)
library(ggplot2)

coefs <- as.matrix(coef(ridge$finalModel, s = ridge$bestTune$lambda))
coef_df <- tibble(term = rownames(coefs), beta = as.numeric(coefs)) %>%
  filter(term != "(Intercept)") %>%
  mutate(abs_beta = abs(beta)) %>%
  arrange(desc(abs_beta)) %>%
  slice_head(n = 15)

ggplot(coef_df, aes(x = reorder(term, abs_beta), y = beta)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top 15 predictors by |β| (ridge)",
       x = "Predictor", y = "Coefficient (signed)") +
  theme_minimal()

```

```{r}
library(dplyr)

# Safety checks (optional but helpful)
stopifnot(exists("R2_base"), exists("MAE_base"), exists("RMSE_base"))
stopifnot(exists("rsq_te"),  exists("mae_te"),  exists("rmse_te"))

# Build comparison table
comp <- bind_rows(
  data.frame(Model = "Mean Only", R2 = R2_base, MAE = MAE_base, RMSE = RMSE_base),
  data.frame(Model = "Ridge",     R2 = rsq_te,  MAE = mae_te,  RMSE = rmse_te)
)

# Round numeric columns
comp_rounded <- comp %>% mutate(across(where(is.numeric), ~ round(.x, 3)))
comp_rounded

```

```{r}
# ✅ Step 8: Compare to baselinec
# dplyr way
library(dplyr)
comp_rounded <- comp %>% mutate(across(where(is.numeric), ~ round(.x, 3)))
comp_rounded
```

```{r}
print(comp, digits = 3)
```

## Model 2: Lasso Regression Model

```{r}
set.seed(20251102)
autism_tr
autism_te
```

```{r}
# ✅ LASSO using same recipe and CV folds
grid <- data.frame(alpha = 1, lambda = seq(0.01, 3, 0.01))

lasso <- caret::train(
  blueprint,
  data      = autism_tr,
  method    = "glmnet",
  trControl = cv,
  tuneGrid  = grid
)
```

```{r}
plot(lasso)           # CV curve
lasso$bestTune        # chosen lambda
lasso$results[1:6, ]  # peek at results
```

```{r}
pred_te_lasso <- predict(lasso, newdata = autism_te)

rsq_te_lasso  <- cor(autism_te$score, pred_te_lasso, use = "complete.obs")^2
mae_te_lasso  <- mean(abs(autism_te$score - pred_te_lasso))
rmse_te_lasso <- sqrt(mean((autism_te$score - pred_te_lasso)^2))

cat("LASSO test R²:", round(rsq_te_lasso, 3),
    "\nMAE:", round(mae_te_lasso, 3),
    "\nRMSE:", round(rmse_te_lasso, 3), "\n")
```

```{r}

coefs <- as.matrix(coef(lasso$finalModel, s = lasso$bestTune$lambda))
coef_df <- tibble(term = rownames(coefs), beta = as.numeric(coefs)) %>%
  filter(term != "(Intercept)") %>%
  mutate(abs_beta = abs(beta)) %>%
  arrange(desc(abs_beta)) %>%
  slice_head(n = 15)

ggplot(coef_df, aes(x = reorder(term, abs_beta), y = beta)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top 15 predictors by |β| (LASSO)",
       x = "Predictor", y = "Coefficient (signed)") +
  theme_minimal()
```

1.  LASSO behaves differently by design

    1.  Ridge (α=0) shrinks all coefficients but **keeps** them non-zero.

    2.  LASSO (α=1) shrinks many to **exactly zero** and often picks **one variable from a correlated group** as the proxy.\

        That’s why your LASSO plot shows one bar (e.g., Class.ASD_NO) dominating while ridge spreads weight across several.

2.  Different λ chosen

    1.  `ridge$bestTune$lambda` and `lasso$bestTune$lambda` are usually different.

    2.  A larger λ with LASSO ⇒ more zeros ⇒ one/few big bars.

3.  Dummy coding & reference levels

    1.  Factors become dummies. With LASSO, often only **one level’s dummy** survives; ridge can assign small weights to several levels

    If `Class.ASD_binary` wasn’t removed (or was removed in ridge but not in lasso), that can also explain differences.

```{r}
comp <- rbind(
  data.frame(Model = "Mean Only", R2 = R2_base, MAE = MAE_base, RMSE = RMSE_base),
  data.frame(Model = "Ridge",     R2 = rsq_te,  MAE = mae_te,  RMSE = rmse_te),
  data.frame(Model = "Lasso",     R2 = rsq_te_lasso, MAE = mae_te_lasso, RMSE = rmse_te_lasso)
)

comp_rounded <- comp %>% mutate(across(where(is.numeric), round, 3))
comp_rounded
```

## Model 3: Elastic Net Model

```{r}
set.seed(20251102)  # reproducibility

loc        <- sample(seq_len(nrow(autism_final)), round(nrow(autism_final) * 0.9))
autism_tr  <- autism_final[loc, ]
autism_te  <- autism_final[-loc, ]

dim(autism_tr)
dim(autism_te)
```

```{r}
blueprint <- recipe(score ~ ., data = autism_tr) %>%
  step_rm(matches("^A(10|[1-9])_Score$"), Class.ASD_binary) %>%  # drop A1_Score…A10_Score + ASD_binary
  step_zv(all_numeric()) %>%
  step_nzv(all_numeric()) %>%
  step_normalize(all_numeric_predictors())
```

```{r}
# Randomly shuffle the training set
autism_tr <- autism_tr[sample(nrow(autism_tr)), ]

# 10 folds of ~equal size
folds <- cut(seq(1, nrow(autism_tr)), breaks = 10, labels = FALSE)

# Indices list for caret
my.indices <- vector("list", 10)
for (i in 1:10) {
  my.indices[[i]] <- which(folds != i)
}

cv <- trainControl(method = "cv", index = my.indices)

caret::getModelInfo()$glmnet$parameters
```

```{r}
grid <- expand.grid(
  alpha  = seq(0, 1, 0.01),     # ridge → lasso
  lambda = seq(0.001, 0.5, 0.005)
)
grid
```

```{r}
Sys.time()

elastic <- caret::train(
  blueprint,
  data      = autism_tr,
  method    = "glmnet",
  trControl = cv,
  tuneGrid  = grid
)

Sys.time()
```

```{r}
plot(elastic)

elastic$bestTune        # chosen alpha & lambda
head(elastic$results)   # peek first rows
# Example: elastic$results[422, ]  # if you want to inspect a specific row

```

```{r}
options(scipen = 99)

coefs <- coef(elastic$finalModel, elastic$bestTune$lambda)
coefs

# Count exact zeros
coefs.zero <- coefs[which(coefs[, 1] == 0), ]
length(coefs.zero)

# Rank by absolute magnitude (exclude intercept)
ind <- order(abs(coefs), decreasing = TRUE)
head(as.matrix(coefs[ind[-1], ]), 10)

```

```{r}
require(vip)

vip(elastic, num_features = 10, geom = "point") +
  theme_bw()

```

```{r}
predict_te_elastic <- predict(elastic, autism_te)
predict_te_elastic

rsq_te  <- cor(autism_te$score, predict_te_elastic, use = "complete.obs")^2
rsq_te

mae_te  <- mean(abs(autism_te$score - predict_te_elastic))
mae_te

rmse_te <- sqrt(mean((autism_te$score - predict_te_elastic)^2))
rmse_te

```

```{r}
library(dplyr); library(tibble); library(ggplot2)

coefs_en <- as.matrix(coef(elastic$finalModel, s = elastic$bestTune$lambda))
coef_df_en <- tibble(term = rownames(coefs_en), beta = as.numeric(coefs_en)) %>%
  filter(term != "(Intercept)") %>%
  mutate(abs_beta = abs(beta)) %>%
  arrange(desc(abs_beta)) %>%
  slice_head(n = 15)

ggplot(coef_df_en, aes(x = reorder(term, abs_beta), y = beta)) +
  geom_col() + coord_flip() +
  labs(
    title = paste0("Top 15 predictors by |β| (Elastic Net, α=",
                   round(elastic$bestTune$alpha, 2), ", λ=",
                   round(elastic$bestTune$lambda, 3), ")"),
    x = "Predictor", y = "Coefficient (signed)"
  ) +
  theme_minimal()
```

```{r}
# Rename your Elastic Net metrics to the expected names
rsq_te_enet  <- rsq_te
mae_te_enet  <- mae_te
rmse_te_enet <- rmse_te
```

```{}
```

```{r}
comp <- rbind(
  data.frame(Model = "Mean Only",  R2 = R2_base, MAE = MAE_base, RMSE = RMSE_base),
  data.frame(Model = "Ridge",      R2 = rsq_te,  MAE = mae_te,  RMSE = rmse_te),
  data.frame(Model = "Lasso",      R2 = rsq_te_lasso, MAE = mae_te_lasso, RMSE = rmse_te_lasso),
  data.frame(Model = "ElasticNet", R2 = rsq_te_enet,  MAE = mae_te_enet,  RMSE = rmse_te_enet)
)

comp_rounded <- comp %>%
  dplyr::mutate(across(where(is.numeric), round, 3))

comp_rounded
```

```{r}
cat("Elastic Net R²:", round(rsq_te_enet, 3),
    "\nMAE:", round(mae_te_enet, 3),
    "\nRMSE:", round(rmse_te_enet, 3), "\n")

```

## Specific setting for model fitting

```{r}
```

## Evaluating model performance

```{r}

```

## Model fit - Model performance

```{r}

```

## Model fit - Final model selection

```{r}

```

## Model fit - Cut off point

```{r}

```

## Model fit - Other considerations

```{r}

```

## Data visualization 1

```{r}

```

## Data visualization 2

```{r}

```
